Copulas and their potential for ecology

### 2019-09-11 Paper.Rmd
still working on Background

<!--***DANr: Add a new section called Background on copulas.
It can have the following kinds of stuff (probably not exhaustive, and probably
not in order, and some of this material is currently below):
1) DONE - The definition, Sklar's theorem
2) DONE What good are copulas - intuitive idea, basically, but also have to be linked to 
the formalities of the definition
3) What some of the different copula families look like - in particular you will use BB1 in Fig S2 and cite Background, so mention it
  Three new figures: Normal, Clayton, BB1
4) What tail dependence actually is

@@

We introduce some standard copula families, as example illustran

We introduce the copula families and 
measures of tail dependence we used.
The normal copula family (henceforth N) is a one-parameter
family, with parameter corresponding to the correlation of the related 
bivariate normal distribtion, and controlling the degree of dependence 
(Fig. \ref{fig_cop_pedag1}G for an example).
Left- and right-tail dependence are $0$ for the N copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measures correspond to dependence stronger than an N
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}C-D) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}C) and $0$ upper-tail dependence; the 
survival-Clayton is a rotation of the Clayton by $180^\circ$ and has 
positive upper-tail ($0.87$ in Fig. \ref{fig_cop_pedag1}D)
and $0$ lower-tail dependence. For our model selection procedure (see below), we used 
several families of copulas exhibiting various 
lower- and upper-tail dependence characteristics 
(Clayton, survival-Clayton, Gumbel, survival-Gumbel, Joe, survival-Joe, Frank,
BB1, survival-BB1,
BB6, survival-BB6, BB7, survival-BB7, BB8, survival-BB8, henceforth
abbreviated C, SC, G, SG, J, SJ, F, BB1, SBB1, BB6, SBB6, BB7, SBB7, BB8, SBB8).
Copulas are implemented in the R packages `copula` and `VineCopula`. 
Visualizations of these families (Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8})
can be compared with normalized rank plots to estimate visually 
which families may effectively model a given bivariate dataset. 
If LT and UT denote 
lower- and upper-tail dependence measures, their difference measures asymmetry of 
tail dependence.
See @nelsen2006_copula, @joe2014_dependence for further background.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7.5in]{./Results/PedagogSuppMat_Normal.pdf}
\caption[Example N copulas]{Log-transformed pdfs (A-E) and samples (F-J) from example N copulas. 
S is Spearman correlation; p is the value of the parameter for the N 
family (it is a 
one-parameter family); and LT and UT are the measures of lower- and upper-tail dependence, respectively (see Background section for definition).\label{figped_N}}
\end{center}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7.5in]{./Results/PedagogSuppMat_Clayton.pdf}
\caption[Example C copulas]{Log-transformed pdfs (A-E) and samples (F-J) from example C 
copulas. S is Spearman correlation; p is the value of the parameter for the C family (it 
is a one-parameter family); and LT and UT are the measures of lower- and 
upper-tail dependence, 
respectively (see Background section for definition).\label{figped_C}}
\end{center}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7.5in]{./Results/PedagogSuppMat_BB1.pdf}
\caption[Example BB1 copulas]{Log-transformed pdfs for example BB1 copulas. K is Kendall correlation; p1 and p2
denote the two parameters of the family (it is a 
two-parameter family); and LT and UT are our measures of lower- and upper-tail dependence, 
respectively (see Background section for definition).\label{figped_BB1}}
\end{center}
\end{figure}

>>>>

We introduce some standard copula families, as examples of copulas and also
because we will fit these families to data later. The bivariate normal
copula family (henceforth abbreviated N), which consists of copulas of bivariate normal
distributions, is a one-parameter family. The parameter, which we here call $p$,
corresponds to the correlation
of the related bivariate normal distribution, and controls the degree of dependence
between the variables. An N copula with $p=0.7$ was already introduced (Fig.
\ref{PedagogFigCopThry}A). The formula for the
copula is $\Phi_{2,p}(\Phi_{1}^{-1}(u),\Phi_{1}^{-1}(v))$, where $\Phi_1^{-1}$ is the inverse of the
cdf of a (univariate) standard normal and $\Phi_{2,p}$ is the joint cdf of a
bivariate normal distribution with mean $(0,0)$ and covariance matrix having $1$s on the diagonal
and $p$ in the off diagonal positions. However, formulas for copulas are often not as 
directly informative about copula properties as pdf diagrams; Fig. \ref{figped_N} has 
example pdfs correspondng to bivariate N copulas for various values of $p$.
Note that the pdfs are symmetric
around the diagonal line $v=-u+1$ (Fig. \ref{figped_N}), so normal copulas have symmetric 
tail dependence (see
below for a more precise defintion of tail dependence). 
The Clayton copula family (henceforth abbreviated C),
of which we also already pictured an example (Fig.
\ref{PedagogFigCopThry}B), is another one-parameter family. In contrast to N copulas, C copulas
have stronger left- than right-tail dependence. The formula is 
$\left[  \reumax(u^{-p}+v^{-p}-1,0)\right]^{-1/p}$, for parameter $p$; though this again provides 
less direct intuition than do example pdfs (Fig. \ref{figped_C}). The survival
Clayton family (here abbreviated SC) is the symmetric opposite of the C family, showing
stronger right- than left tail dependence (Fig. \ref{SM-figped_C}). The BB1 copula 
family is a two-parameter family, thereby providing more flexibility: for some parameters
BB1 copulas have stronger left- than right-tail dependence, and for others the reverse
(Fig. \ref{figped_BB1}). See @nelsen2006_copula and @joe2014_dependence for further
information on these and other copula families.

Next intro tail dependence

<!--***DANr: Add a new section called Background on copulas.
It can have the following kinds of stuff (probably not exhaustive, and probably
not in order, and some of this material is currently below):
1) DONE - The definition, Sklar's theorem
2) DONE What good are copulas - intuitive idea, basically, but also have to be linked to 
the formalities of the definition
3) DONE What some of the different copula families look like - in particular you will use BB1 in Fig S2 and cite Background, so mention it
  Three new figures: Normal, Clayton, BB1
4) What tail dependence actually is

@@

<!--***DAN: text storage for later:
Left- and right-tail dependence are $0$ for the N copula, and this 
serves as a benchmark for tail dependence in other copulas - positive values of our
tail dependence measures correspond to dependence stronger than an N
copula. The Clayton and 
survival-Clayton families (Fig. \ref{fig_cop_pedag1}C-D) are two more one-parameter 
families. The Clayton has positive lower-tail dependence ($0.87$ for
the example of Fig. \ref{fig_cop_pedag1}C) and $0$ upper-tail dependence; the 
survival-Clayton is a rotation of the Clayton by $180^\circ$ and has 
positive upper-tail ($0.87$ in Fig. \ref{fig_cop_pedag1}D)
and $0$ lower-tail dependence. -->

<!--***DAN: put this text in the methods. It won't help people here in the ms to have all
these copulas intro'd. The sole purpose of them is give a variety of options to have a robust model
selection and multi-model inference procedure for tail dependence, so mention them in that context in 
the Methods in a way that make clear we don't need to give a damn about these things:
For our model selection procedure (see below), we used 
several families of copulas exhibiting various 
lower- and upper-tail dependence characteristics 
(Clayton, survival-Clayton, Gumbel, survival-Gumbel, Joe, survival-Joe, Frank,
BB1, survival-BB1,
BB6, survival-BB6, BB7, survival-BB7, BB8, survival-BB8, henceforth
abbreviated C, SC, G, SG, J, SJ, F, BB1, SBB1, BB6, SBB6, BB7, SBB7, BB8, SBB8).
Copulas are implemented in the R packages `copula` and `VineCopula`. 
Visualizations of these families (Figs \ref{SM-figped_N}-\ref{SM-figped_SBB8})
can be compared with normalized rank plots to estimate visually 
which families may effectively model a given bivariate dataset. 
If LT and UT denote 
lower- and upper-tail dependence measures, their difference measures asymmetry of 
tail dependence.
See @nelsen2006_copula, @joe2014_dependence for further background. -->

\begin{figure}[!h]
\begin{center}
\includegraphics[width=\textwidth]{./Results/PedagogSuppMat_Normal.pdf}
\caption[Example N copulas]{Log-transformed pdfs (A-E) and samples (F-J) from example N copulas. 
S is Spearman correlation; p is the value of the parameter for the N 
family (it is a 
one-parameter family); and LT and UT are the measures of lower- and upper-tail dependence, respectively (see Background section for definition).\label{figped_N}}
\end{center}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=\textwidth]{./Results/PedagogSuppMat_Clayton.pdf}
\caption[Example C copulas]{Log-transformed pdfs (A-E) and samples (F-J) from example C 
copulas. S is Spearman correlation; p is the value of the parameter for the C family (it 
is a one-parameter family); and LT and UT are the measures of lower- and 
upper-tail dependence, 
respectively (see Background section for definition).\label{figped_C}}
\end{center}
\end{figure}

\begin{figure}[!h]
\begin{center}
\includegraphics[width=\textwidth]{./Results/PedagogSuppMat_BB1.pdf}
\caption[Example BB1 copulas]{Log-transformed pdfs for example BB1 copulas. K is Kendall correlation; p1 and p2
denote the two parameters of the family (it is a 
two-parameter family); and LT and UT are our measures of lower- and upper-tail dependence, 
respectively (see Background section for definition).\label{figped_BB1}}
\end{center}
\end{figure}
